Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-27 introduction.tex
Small cleanups and added Alessandra to acknowledgements
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@760 ccdd05d5-56c1-461a-bf98-85f6876523b4
Updated astrophysical motivation; added many citations to observations.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@765 ccdd05d5-56c1-461a-bf98-85f6876523b4

Prompt electromagnetic emission can arise as shells of relativistically
outflowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts~\cite{Lee:2005, nakar07}. This prompt emission is a probe into the
extreme initial conditions of the outflow. A very bright optical flash should
accompany the x-rays and gamma-rays~\cite{Sari99}. \editorial{Should mention
that optical flashes have been seen for a handful of long GRBs by very fast
slews. TAROT has seen many within ~20 seconds. \cite{Klotz09}} In contrast, an
afterglow is produced when the outflow impacts the interstellar medium.
Afterglows represent lower Lorentz factors and the conditions of the
environment far outside of the central engine, and are relatively insensitive
to initial conditions. In order to observe the most intense electromagnetic
emission and provide an unbroken chain of observations between the progenitor
system and the expanding fireball, telescopes must point during the seconds
surrounding the merger.

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \textsc{ligo} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current trigger generation and telescope actuation would then dominate
latency.

To this end, we have the ambition of reporting candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  \editorial{I think we
should say also something about how we can also get the full SNR within moments
after the coalescence, it is obvious maybe, but we should be explicit}  By
looking for threshold crossings before the signal leaves the detection band, it
is possible to trade some signal to noise ratio (\textsc{snr}) for latency.
Figure \ref{fig:earlywarning} shows projected early trigger rates for NS--NS
binaries in Advanced \LIGO\ assuming the event rate predictions in
\cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
bandlimited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with predictive detection of compact binary sources is
possible with current computing resources.  Assuming other aspects of
gravitational-wave observation latency can be reduced significantly, this
should allow the possibility for prompt alerts to be sent to the astronomical
community.

>>>>

Prompt electromagnetic emission can arise as shells of relativistically
outflowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short GRBs)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which are
relatively insensitive to initial conditions. Optical flashes
have only been observed for a handful of long
GRBs~\cite{1999Natur.398..400A,2003Natur.422..284F,2006A&A...454L.119J,0004-637X-660-1-489,2008Natur.455..183R,2011A&A...528A..15G}
by telescopes with extremely rapid response or, in the case of GRB 080319B, by
pure serendipity, where several telescopes were observing a previous GRB in the
same field~\cite{2008Natur.455..183R}. Short GRBs, on the other hand, typically
fade too quickly to observe past the initial burst of gamma rays and hard
x-rays. Rapid gravitational-wave transient alerts could enable the observation
of optical flashes from short GRBs. An optical counterpart would vastly boost
confidence in the gravitational-wave detection and provide the tight sky
localization necessary to allow determination of the source's host galaxy,
which leads to a redshift measurement. With both redshift and a coincident
gravitational-wave observation, we can produce precision measurements of the
Hubble constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  \editorial{I think we
should say also something about how we can also get the full \SNR{} within moments
after the coalescence, it is obvious maybe, but we should be explicit}  By
looking for threshold crossings before the gravitational-wave signal leaves
the detection band, it is possible to trade some signal to noise ratio (\SNR{})
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for NS--NS binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources detectable within 10
seconds of merger and $\sim1$ sources detectable within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above \SNR{} 8.}
\end{figure}

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current trigger generation and telescope actuation would then dominate
latency.

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
bandlimited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank-reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with predictive detection of compact binary sources is
possible with current computing resources.  Assuming other aspects of
gravitational-wave observation latency can be reduced significantly, this
should allow the possibility for prompt alerts to be sent to the astronomical
community.
